---
title: "DGE analysis of SLCs"
output: 
  html_document: 
    df_print: paged
---

```{r}
library(tidyverse)
#library(stringr)
#library(dplyr)
library(limma)
library(edgeR)
library(ggplot2)
library(ggdendro)
library(pvclust)
library(gplots)
library(cluster)
library(igraph)
```
First we copy the [list of SLCs in mammalian genomes](https://esbl.nhlbi.nih.gov/Databases/SLC-list/) and convert gene IDs to ENSG IDs with tool on this [page](https://biit.cs.ut.ee/gprofiler/convert).Then we can select all SLCs in GTEx gene counts data file with `filter.py` in terminal.
```{bash, eval=FALSE}
python filter.py -f1 ../Input/gProfiler_hsapiens_12-11-2023_6-02-56\ AM.csv -f2 ../Input/bulk-gex_v8_rna-seq_GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_reads.gct -o ../Input/GTEx_SLCs_gene_reads.gct
```

```{r}
# Load count data
counts.data <- read_tsv("../Input/GTEx_SLCs_gene_reads.gct")
samp.anno.data <- read_tsv("../Input/GTEx_Analysis_v8_Annotations_SampleAttributesDS.txt")
subj.anno.ds.data <- read_tsv("../Input/GTEx_Analysis_v8_Annotations_SubjectPhenotypesDS.txt")
counts.data <- counts.data[,-1]

# Retrieve the count data of 2 among so many various tissues to analyse. Subsequent analysis should include all the data
tiss1.sampid <- samp.anno.data %>%
  filter(SMTS=='Ovary') %>% 
  select(SAMPID)

tiss2.sampid <- samp.anno.data %>% 
  filter(SMTS=='Testis') %>% 
  select(SAMPID)


ids <- tiss1.sampid %>% 
  bind_rows(tiss2.sampid) %>% 
  add_row(SAMPID='Description') %>% 
  pull() %>% 
  paste(collapse="|")

counts.data <- counts.data %>% 
  select(matches(ids))

# For our subsequent analyses we want to reduce the data set to only those genes with some expression. In this case we will discard genes with no reads in >= 50 samples
counts.data <- counts.data[rowSums(counts.data[,-1]==0)<50,]

#cor.table <- cor(counts.data[,-1])
#cor.table %>% gplots::heatmap.2(dendrogram="row", trace = "none", col=viridis::viridis(25, begin=.25), margins=c(7,8))

# Create a tibble for edgeR description
tiss1.sampid <- tiss1.sampid %>% 
  mutate(TS='Ovary')

tiss2.sampid <- tiss2.sampid %>% 
  mutate(TS='Testis')
```
Before we can normalize the data we need to be able to tell edgeR which groups the samples belong to. Here group refers to the distinct sample types (i.e. combinations of tissue, age, death circumstance)
```{r}
sample.description <- tiss1.sampid %>% 
  bind_rows(tiss2.sampid) %>%
  mutate(SUBJID=str_extract(SAMPID, 'GTEX-.+(?=-.+-SM)')) %>%
  left_join(subj.anno.ds.data, by='SUBJID') %>% 
  mutate(group=paste(TS, SEX, AGE, DTHHRDY, sep="_")) %>% 
  select(-SUBJID)

sample.description <- sample.description %>%
  mutate(TS=factor(TS,levels = c("Testis","Ovary")), # setting the levels in this way makes "Testis" the reference  
         SEX=factor(SEX,levels = c(1,2)),
         AGE=factor(AGE,levels = c("20-29","30-39","40-49","50-59","60-69","70-79")),
         DTHHRDY=factor(DTHHRDY,levels = c(0,1,2,3,4)))

sample.description
```
Calculate normalization factors by TMM method while RPKM is poorly behaved statistically, and plot multidimensional scaling plot of distances between gene expression profiles by logFC methods instead of bcv due to the large quantity of samples
```{r}
counts.matrix <- counts.data %>% select(-Description) %>% as.matrix()
rownames(counts.matrix) <- counts.data$Description

sample.description <- tibble(SAMPID=colnames(counts.matrix)) %>%
  left_join(sample.description)

dge.data <- DGEList(counts=counts.matrix, 
                    group=sample.description$group)
 
dge.data <- calcNormFactors(dge.data, method = "TMM")
dge.data$samples

# Make a plot of the Biological Coefficient of Variation of each sample

# Try to find out if sample can be divided by variables (i.e. TS, AGE, SEX)
plot_info <- sample.description %>% 
  mutate(col=ifelse(TS=="Testis","#000000","#FFFFFF"),
         pch=ifelse(SEX==1,21,24),
         bg=case_when(AGE=="20-29"~"#1F77B4",
                       AGE=="30-39"~"#FF7F0E",
                       AGE=="40-49"~"#2CA02C",
                       AGE=="50-59"~"#D62728",
                       AGE=="60-69"~"#9467BD",
                       AGE=="70-79"~"#8C564B")
    ) %>% 
  select(SAMPID,bg,pch,col)

# (i.e. TS, AGE, SEX, DTHHRDY)
#plot_info <- sample.description %>% 
#  mutate(col=ifelse(TS=="Testis","#000000","#FFFFFF"),
#         pch=ifelse(SEX==1,21,24),
#         bg=case_when(DTHHRDY==0~"#1F77B4",
#                       DTHHRDY==1~"#FF7F0E",
#                       DTHHRDY==2~"#2CA02C",
#                       DTHHRDY==3~"#D62728",
#                       DTHHRDY==4~"#9467BD",
#                       TRUE~"#8C564B")
#    ) %>% 
#  select(SAMPID,bg,pch,col)

matched_bgs <- plot_info$bg[match(colnames(dge.data), plot_info$SAMPID)]
matched_pchs <- plot_info$pch[match(colnames(dge.data), plot_info$SAMPID)]
matched_cols <- plot_info$col[match(colnames(dge.data), plot_info$SAMPID)]

# Unique age groups and corresponding colors
age_groups <- c("20-29", "30-39", "40-49", "50-59", "60-69", "70-79")
age_colors <- c("#1F77B4", "#FF7F0E", "#2CA02C", "#D62728", "#9467BD", "#8C564B")

# Sex symbols
sex_symbols <- c(21, 24)
sex_labels <- c("Male", "Female")

# Tissue outline colors
ts_groups <- c("Testis", "Ovary")
ts_colors <- c("#000000","#FFFFFF")


plotMDS(dge.data,method="logFC",pch=matched_pchs,bg=matched_bgs,col=matched_cols,
        cex=1.4, dim.plot=c(1,2))

# Add a legend for Age groups
legend("bottom",  # Places the legend in the top right corner of the plot
       legend = age_groups,
       fill = age_colors,
       title = "Age",
       cex = 0.8,
       y.intersp = 0.7)  # Adjusts the size of the text in the legend

# Add another legend for Sex
legend("topleft",  # Places the legend in the bottom right corner of the plot
       legend = sex_labels,
       pch = sex_symbols,
       title = "Sex",
       cex = 0.8,
       y.intersp = 0.7)
# Add the last for TS
legend("topright",  # Places the legend in the top left corner of the plot
       legend = ts_groups,
       col = ts_colors,
       border = ts_colors,
       fill = "#FFFFFF",
       pch = 21,
       title = "Tissue",
       cex = 0.8,
       y.intersp = 0.7)  # Adjusts the size of the text in the legend

#plotMDS(dge.data,method="logFC",pch=matched_pchs,bg=matched_bgs,col=matched_cols,
#        cex=1.4, dim.plot=c(3,2))

```
We can see samples are well divided into 2 clusters by tissue. This plot could convince us of the correctness of the experiment and data.
Randomly select 6 samples and plot them to check the effect of normalization
```{r}
counts.data.normal <- cpm(dge.data) 

# or log2 transformed:
counts.data.normal.log <- cpm(dge.data,log = TRUE)

counts.data.log <- log2(counts.data[,-1] + 1)

counts.data.log.pvt <- counts.data.log

counts.data.log.pvt <- counts.data.log.pvt %>% 
  pivot_longer(colnames(counts.data.log.pvt), names_to="sample", values_to="value")

counts.data.log.pvt <- sample.description %>% 
  select(SAMPID, group) %>% 
  right_join(counts.data.log.pvt, by=join_by(SAMPID==sample))

counts.data.log.pvt %>% ggplot(aes(x=group, y=value)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 70))

counts.data.normal.log.pvt <- as.data.frame(counts.data.normal.log)
  

counts.data.normal.log.pvt <- counts.data.normal.log.pvt %>% 
  pivot_longer(colnames(counts.data.normal.log.pvt), names_to="sample", values_to="value")

counts.data.normal.log.pvt <- sample.description %>% 
  select(SAMPID, group) %>% 
  right_join(counts.data.normal.log.pvt, by=join_by(SAMPID==sample))

counts.data.normal.log.pvt %>% ggplot(aes(x=group, y=value)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 70))
```

```{r}
# Now calculate dispersion factors
design <- model.matrix(~TS,data = sample.description)
rownames(design) <- sample.description$SAMPID

# First the overall dispersion
dge.data <- estimateGLMCommonDisp(dge.data,design,verbose = TRUE)

# Then a trended dispersion based on count level
dge.data <- estimateGLMTrendedDisp(dge.data,design)

# And lastly we calculate the gene-wise dispersion, using the prior estimates to 
#"squeeze" the dispersion towards the common dispersion.
dge.data <- estimateGLMTagwiseDisp(dge.data,design)
```

```{r}
# Draw dispersion plot
plotBCV(dge.data)
```

```{r}
# Find genes that expressed differently in these 2 tissues
fit <- glmFit(dge.data, design)

ts.lrt <- glmLRT(fit,coef = "TSOvary")

# label genes with a FDR < 0.05 and a fold change > 1.5 and draw a volcano plot
results <- topTags(ts.lrt, n=Inf)$table
results$significant <- results$FDR < 0.05 & abs(results$logFC) > log2(1.5)
results$PValue <- pmax(results$PValue, 1e-300)  # replace exact 0s with a very small number

results$SignificanceLabel <- ifelse(results$significant, "Significant", "Not Significant")

ggplot(results, aes(x = logFC, y = -log10(PValue), col=significant)) +
  geom_point(alpha = 0.5) +
  geom_vline(xintercept = c(-log2(1.5),log2(1.5)), linetype="dashed",color="blue") +  # LogFC lines
  geom_hline(yintercept = -log10(0.05),linetype = "dashed",color = "blue") +  # FDR lin
  theme_minimal() +
  labs(x="Log2 Fold Change", y="-Log10 Adjusted P-value", title = "Volcano Plot")
# Summarize the number of differentially expressed genes
summary(decideTestsDGE(ts.lrt,p.value=0.05,lfc=log2(1.5)))

## Go enrichment analysis left to do here

```



```{r}
## As Most of the simplest clustering assumes data to be from a normal 
##distribution. we need to transformed the RNAseq data to be normally 
##distributed

v <- voom(counts.matrix, design, plot=TRUE)
voom_E <- as.data.frame(v$E) %>% 
  rownames_to_column(var = "GeneID")

voom_Et <- voom_E %>%
  t() %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "SAMPID") %>% 
  left_join(sample.description,by="SAMPID")

colnames(voom_Et) <- c("SAMPID",voom_Et[1,2:333],1,2,3,4,"group")
voom_Et <- voom_Et[-1,-334:-337] %>% 
  mutate(across(c(2:333), as.numeric))

#voom_E

mean_vals <- voom_Et %>%
  group_by(group) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE))
 
mean_vals_t <- mean_vals %>% 
  t()
colnames(mean_vals_t) <- mean_vals_t[1,]

mean_vals_t <- mean_vals_t[-1,] %>% 
  as.data.frame() %>% 
  mutate(across(c(1:48), ~as.numeric(.)))

mean_vals_t

E_matrix <- mean_vals_t %>%
  as.matrix() %>% 
  scale() # scale and center the data so that each gene has a mean of 0 and a standard deviation of 1. This prevents genes with high expression from having an undue influence on our results
gene_hclust_row <- E_matrix %>% dist() %>% hclust()

gene_hclust_col <- E_matrix %>% t()  %>% dist() %>% hclust()
ggdendrogram(gene_hclust_col)

#set.seed(12456) #This ensure that we will have consistent results with one another

#fit <- pvclust(E_matrix, method.hclust = "ward.D", method.dist = "euclidean", nboot = 100)

#plot(fit, print.num=FALSE) # dendogram with p-values

heatmap.2(E_matrix, Rowv = as.dendrogram(gene_hclust_col),  density.info="none", 
          trace="none", margins = c(20,10))

```


```{r}
# K-means section

## get principle components
prcomp_counts <- prcomp(E_matrix) 
scores <- as.data.frame(prcomp_counts$x)[,c(1,2)]

## Get appropriate cluster number
#set.seed(125)
gap <- clusGap(E_matrix, FUN=kmeans, iter.max=10000, K.max=30, B=100)

plot(gap, main = "Gap Statistic")

with(gap, maxSE(Tab[,"gap"], Tab[,"SE.sim"], method="firstSEmax"))

# We now try 5 clusters
set.seed(25)
fit <- kmeans(E_matrix, 4)
clus <- as.data.frame(fit$cluster)
names(clus) <- paste("cluster")

plotting <- merge(clus, scores, by = "row.names")
plotting$cluster <- as.factor(plotting$cluster)

# plot of observations
ggplot(data = plotting, aes(x = PC1, y = PC2, label = Row.names, color = cluster)) +
  geom_hline(yintercept = 0, colour = "gray65") +
  geom_vline(xintercept = 0, colour = "gray65") +
  geom_point(alpha = 0.8, size = 2, stat = "identity") 
```

```{r}
clus <- cbind(clus, E_matrix) %>% # add cluster labels to gene expression matrix
  mutate(gene_index=1:nrow(clus)) # would be better to bring in gene names but OK for now.

clus <- clus %>% 
  pivot_longer(c(-cluster, -gene_index), names_to = "sample_ID", values_to = "expression") %>% # get ready for plotting
  mutate("sample_group" = str_remove(sample_ID, "(_[0134]|NA)$"))

clus_summary <- clus %>% # average across replicates
  group_by(gene_index, cluster, sample_group) %>%
  summarize(expression = mean(expression))

clus_summary %>% ggplot(aes(x=sample_group, y=expression, group=gene_index)) + # plot it
  geom_line(alpha=.15) +
  facet_wrap(~cluster, ncol=4) +
  coord_fixed(ratio=1) +
  theme(axis.text.x = element_text(angle = 90, size = 7, hjust = 1, vjust = 0))
```
```{r}
#  Co-Expression
## Get DGEs
DEgenes <- topTags(ts.lrt,n = Inf,p.value = 0.05)$table
voom_E_ts <- voom_E %>%
  filter(GeneID %in% rownames(DEgenes))

## Turn to matrix
E_matrix_ts <- voom_E_ts %>%
  as.data.frame() %>%
  column_to_rownames("GeneID") %>%
  as.matrix()

## Calculate mutual ranks
E_matrix_cor <- cor(t(E_matrix_ts))
diag(E_matrix_cor) <- 0
E_matrix_rank <- apply(E_matrix_cor,2,function(x) rank(-abs(x)))
E_matrix_MR <- sqrt(E_matrix_rank * t(E_matrix_rank))

genes_adj_MR4 <- E_matrix_MR <= 4
diag(genes_adj_MR4) <- 0

genes_adj_MR10 <- E_matrix_MR <= 10
diag(genes_adj_MR10) <- 0

## Plot our networks using different thresholds for connectivity
gene_graphMR4 <- graph.adjacency(genes_adj_MR4, mode = "undirected") #convert adjacency to graph
compsMR4 <- clusters(gene_graphMR4)$membership                        #define gene cluster membership
colbar <- rainbow(max(compsMR4)+1)                                   #define colors
V(gene_graphMR4)$color <- colbar[compsMR4+1]                          #assign colors to nodes
plot(gene_graphMR4, layout = layout_with_fr, vertex.size = 4, vertex.label = NA, main="MR 4")
  
gene_graphMR10 <- graph.adjacency(genes_adj_MR10, mode = "undirected") #convert adjacency to graph
compsMR10 <- clusters(gene_graphMR10)$membership                        #define gene cluster membership
colbar <- rainbow(max(compsMR10)+1)                                   #define colors
V(gene_graphMR10)$color <- colbar[compsMR10+1]                          #assign colors to nodes
plot(gene_graphMR10, layout = layout_with_fr, vertex.size = 4, vertex.label = NA, main="MR 10")
```

```{r, include=FALSE, eval=FALSE}
E_matrix_5 <- E_matrix_ts[11:15,]
E_matrix_5_cor <- cor(t(E_matrix_5))

diag(E_matrix_5_cor) <- 0


E_matrix_5_rank <- apply(E_matrix_5_cor,2,function(x) rank(-abs(x)))
E_matrix_5_rank

E_matrix_5_MR <- sqrt(E_matrix_5_rank * t(E_matrix_5_rank))
E_matrix_5_MR %>% round(3)

genes_adj_MR2 <- E_matrix_5_MR <= 2
diag(genes_adj_MR2) <- 0
genes_adj_MR2
```

```{r}
# Graph stats for network comparison
graph.density(gene_graphMR4)
average.path.length(gene_graphMR4)

graph.density(gene_graphMR10)
average.path.length(gene_graphMR10)
```
The decrease of average path length can be caused by the dramatic increase of number of dots that been connected when the threshold grows from 4 to 10. These new connected dots are far away from each other compare to those that has already been connected when MR is 4. This may imply that MR4 is better than MR10 because those loosely connected genes suggest nothing regarding biological connection.

```{r}
# The most central genes
## Calculate degree centrality
colSums(genes_adj_MR4) %>%
  sort(decreasing = TRUE) %>%
  tibble(GeneID=names(.), degree=.) %>% 
  head()
```


